\sys: Property-Guided Program Regeneration\\{\vspace{1em}\large Technical University of Denmark, MSc Thesis {\\\vspace{-1em}} Advisors: Christian Gram Kahlauge and Nikos Vasilakis

### 2025-07-26 docs/paper/draft.tex
Fixes throughout
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{wrapfigure}[4]{r}{.7\columnwidth}
\vspace{-10pt}
\hspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
fsignt(p_, S) :- fsignt(p, S).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys query language to express 
this regeneration goal.
This query requires the regenerated program \ttt{p\_} to exhibit the same I/O behavior as the original program \ttt{p}
but forbids side-effects.
\sys takes this query and arrives at the set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{$\ttt{graph(p, I, O)}, \ttt{inpt(P, I)}, \ttt{fsignt(P, S)}, \ttt{pure(P)}$\}$, where
\ttt{input(p, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\ttt{graph(p, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs
(an example I/O pair is $\langle\ttt{([1,[2,3]], (x)=>x+1)}\to\ttt{[2,3,4]}\rangle$);
and \ttt{fsignt(p, S)} extracts the function's signature using a language-aware parser.
The, \ttt{pure(p)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.
After 30.4 seconds and one attempt, having generated 30 I/O pairs, \sys generates the following program:
\begin{minted}[frame=lines]{js}
module.exports = function(stream, fn) {
  return stream.map(fn); };
\end{minted}

@@

\heading{Planning over properties}
The logic engine determines the set of active properties to enforce by solving
a constrained optimization problem over the extractable property set $P_r$.
It satisfies the
following key constraint:
\begin{align*}
  X &\Leftarrow \mathsf{can}(X) &\text{Apply only extractable properties}
\end{align*}

@@

\heading{Plugin-defined rules}
Plugins can also contribute domain-specific inference rules and constraints.
For example, a plugin for JavaScript-to-Haskell translation may define:
\begin{align*}
\mathsf{can}(\ttt{translate}_{\ttt{js} \rightarrow \ttt{hs}}(p)) & \\
\ttt{translate}_{\ttt{js} \rightarrow \ttt{hs}}(p) &\Rightarrow \ttt{language}(p, \ttt{haskell}) \\
\ttt{translate}_{\ttt{js} \rightarrow \ttt{hs}}(p) &\Rightarrow \mathsf{language}(p, \ttt{javascript})
\end{align*}
Intuitively, these rules mean (1) that the system is able to apply the plugin,
and (2) it can only be applied if the original program is written in JavaScript, 
and the regenerated program \emph{must} be written in Haskell.

@@

\heading{Compilation to ASP}
Given a source program written in the \sys query language, the compiler
performs a syntactic de-sugaring into ASP rules~(\cref{fig:asp-compilation}).
For each rule $H \leftarrow B_1,\dots,B_n$, the compiler first performs a
pseudo-grounding step that replaces every capitalized variable $V_i$ with a
fresh, unique symbol $\mathit{sym}(r,P,V_i)$, generated from the rule
name $r$, target program $P$, and original variable name $V_i$, proceeding
left-to-right to preserve data dependencies.
Each one of these concrete values will correspond to extracted program properties.
After renaming, the head and body
are emitted as an ASP capability rule of the form $\mathit{can}(H') \leftarrow
\mathit{do}(B_1'),\dots,\mathit{do}(B_n')$. 
All implicit query
goals (rule heads $H$) are compiled into hard constraints $\neg \mathit{do}(H) \rightarrow
\bot$, meaning that the ASP program will be unsatisfiable unless all query goals are met.
Optimization directives $\#\min X$ and $\#\max X$ are mapped to
$\mathit{do_min}(X)$ and $\mathit{do_max}(X)$, respectively.
Finally, \sys
appends program-agnostic axioms and adds a choice rules that allows the solver to select which properties to extract from the program in order to satisfy the query.
This consists as the core of the \sys logic program, and the reason the rule to \ttt{can}/\ttt{do} predicates is required.
The result is the compilation is a valid ASP program which is then evaluated by an ASP solver.
The next section describes how the set of produced predicates is used to construct the regeneration plan.

@@

\heading{Logic program resolution}
Once the input query is compiled into an ASP program, \sys invokes an
off-the-shelf ASP solver to compute a stable model that satisfies the query.
The model consists of a set of selected $\mathit{do}$ atoms representing the regeneration plan.
\sys then executes this plan~\cref{alg:resolution}.
First, \sys discards all plugin-activation atoms from the plan, 
as they are only relevant to the logic engine.
Then \sys iterates through each
$\mathit{do}(f(a_1,\dots,a_n))$ atom.
If an argument value is missing, the plugin responsible for producing
it (found by parsing the variable's name) is executed and produced program properties are stored in a value map.
Once all actions have been executed in dependency order, the extracted properties are passed to the regeneration phase.

@@

\heading{Logic and concrete space}
\sys operates at the boundary between two representational domains: a symbolic
\emph{logic space}, where transformations over program properties are expressed, and
a \emph{concrete space}, where actual programs, inputs, outputs, and traces
reside.

In the logic space, properties are encoded as predicates over abstract program
identifiers.
To bridge this abstraction gap, \sys given the set of true predicates
from the logic space, extracts the corresponding concrete properties from the
source program.
For example, the predicate $\ttt{inputs}(p, I)$ in the logic space
will be concertized to set of inputs $I$ by invoking the \ttt{input} plugin,
which uses an LLM to generate a set of inputs for the program $p$.
Similarly, the predicate $\ttt{graph}(p, I, O)$ will be concertized to a set of
input-output pairs $(I, O)$ by executing the program $p$ on the inputs $I$ and
recording the outputs $O$.
After regeneration, the predicates over the concrete regenerated program $p\_$
will be checked regarding their relationship to the same properties over program $p$
in the logic space.
% An equality relationship is one where the predicate's grounded version
% $X$ is the same between the two programs, modulo the program identifier~(\eg, $\ttt{graph}(p, i, o) \equiv \ttt{graph}(p\_, i, o)$).
Equality is defined differently for each property, and is enforced by the corresponding plugin.
Similarly, minimization and maximization relationships are defined as
the concrete property value being less-than or greater-than the original program's value, respectively.
Again, the less-than and greater-than operators are defined by the plugin that verifies the property.

@@

As shown in \cref{fig:logic-to-concrete}, this process constitutes a
translation barrier: declarative goals over properties like \ttt{inputs},
\ttt{outputs}, and \ttt{traces} are realized through observation and
instrumentation. Once aligned, the logic engine can reason over these grounded
facts to drive transformation.

@@

\heading{Answer Set Programming}
The \sys prototype compiles each input query to an answer-set programming (ASP) program~\cite{Eiter_2009}.
A brief introduction to the paradigm follows.
ASP \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo} \cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

The syntax is similar to \textit{Prolog} and \textit{Datalog}, with code and data being represented
through logical terms.
Collections of logical terms, as seen in
\cref{lst:terms}, can represent the program and its properties.
Rules, expressed using
the \ttt{:-} operator, define relationships
between atoms.
Choice rules, exemplified in \cref{lst:choice}, allow the ASP
solver to make selections among atoms.
Integrity constraints, as shown in
\cref{lst:constraint}, can restrict invalid answer sets. Optimization
directives, such as \ttt{\#minimize} and \ttt{\#maximize} in
\cref{lst:optimization}, guide the solver to output optimal answer sets,
with respect to a given metric (\eg the number of properties to be extracted from the input program).

@@

For instance, a plan for regenerating a specific function $f$ might include:
\ttt{language}(f, \ttt{haskell}) the function must be written in Haskell, 
\ttt{pure}(f)) the function must be side-effect-free, and 
\ttt{signature}(f, s)) the function must keep the original program's signature $s$.

@@

\section{Property Verification and Feedback Loop}
\label{sec:verification}

@@

\begin{table}[htpb]
  \centering
  \caption{\textbf{Correctness result summary.} Number of tasks passing all tests (when present) and manual inspection.}
  \begin{tabular}{lc}
    \toprule
    Benchmark & Correctness \\ 
    \midrule
    Rosetta Code & 16/16  \\
    Short Utilities & 10/13 \\
    SSCA & 3/3  \\
    IOCCC & 1/2 \\
    Microbenchmarks & 2.5/3  \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\heading{Idea of program projection}
\sys can be viewed as a projection system, as shown in \cref{fig:projection}.
\sys projects a program into an abstract space of properties, 
then, it performs transformations within that space (the input query),
and finally, it regenerates this projection back into a concrete program.
The idea of casting programs into a high-dimensioanl space has been studied before~\cite{ben-nun2018neural,alon2019code2vec,ellis2020dreamcoder,nye2021platypus},
but \sys's approach, from this perspective, might offer interesting insights.

>>>>

\begin{wrapfigure}[3]{r}{.7\columnwidth}
\vspace{-10pt}
\hspace{-10pt}
\begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
fsignt(p_, S) :- fsignt(p, S).
pure(p_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys query language to express 
this regeneration goal.
This query requires the regenerated program \ttt{p\_} to exhibit the same I/O behavior as the original program \ttt{p}
but forbids side-effects.
\sys takes this query and arrives at the set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{$\ttt{graph(p, I, O)}, \ttt{inpt(P, I)}, \ttt{fsignt(P, S)}, \ttt{pure(P)}$\}$, where
\ttt{input(p, I)} means that \sys will use the \ttt{input} plugin extract a set of inputs from the original program
by providing the program's source code to an LLM instance and asking it to generate a set of inputs;
\ttt{graph(p, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs
(an example I/O pair is $\langle\ttt{([1,[2,3]], (x)=>x+1)}\to\ttt{[2,3,4]}\rangle$);
and \ttt{fsignt(p, S)} extracts the function's signature using a language-aware parser.
The, \ttt{pure(p)} property means that \sys will not provide to the LLM 
instance any information that could encode side-effects, such as the
program's source code, a system call trace, or a filesystem access trace.
After 30.4 seconds and one attempt, having generated 30 I/O pairs, \sys generates the following program:
\begin{minted}[frame=lines]{js}
module.exports = function(stream, fn) {
  return stream.map(fn); };
\end{minted}

@@

\heading{Planning over properties}
The logic engine determines the set of active properties to enforce by solving
a constrained optimization problem over the extractable property set $P_r$.
It satisfies the
following key constraint:
  $X \Leftarrow \mathsf{can}(X)$, apply only extractable properties.

@@

\heading{Plugin-defined rules}
Plugins can also contribute domain-specific inference rules and constraints.
For example, a plugin for JavaScript-to-Haskell translation may define:
\begin{align*}
\mathsf{can}(\ttt{js2hs}(p)) & \\
\ttt{js2hs}(p) &\Rightarrow \ttt{language}(p, \ttt{haskell}) \\
\ttt{js2hs}(p) &\Rightarrow \mathsf{language}(p, \ttt{javascript})
\end{align*}
Intuitively, these rules mean that 
(1) the system is able to apply the plugin,
and (2) it can only be applied if the original program is written in JavaScript, 
and (3) the regenerated program \emph{must} be written in Haskell.

@@

\heading{Compilation to ASP}
Given a source program written in the \sys query language, the compiler
performs a syntactic de-sugaring into ASP rules~(\cref{fig:asp-compilation}).
For each rule $H \leftarrow B_1,\dots,B_n$, the compiler first performs a
pseudo-grounding step that replaces every capitalized variable $V_i$ with a
fresh, unique symbol $\mathit{sym}(r,P,V_i)$, generated from the rule
name $r$, target program $P$, and original variable name $V_i$, proceeding
left-to-right to preserve data dependencies.
Each one of these concrete values will correspond to extracted program properties.
After renaming, the head and body
are emitted as an ASP capability rule of the form $\mathit{can}(H') \leftarrow
\mathit{do}(B_1'),\dots,\mathit{do}(B_n')$. 
All implicit query
goals (rule heads $H$) are compiled into hard constraints $\neg \mathit{do}(H) \rightarrow
\bot$, meaning that the ASP program will be unsatisfiable unless all query goals are met.
Optimization directives $\#\min X$ and $\#\max X$ are mapped to
$\mathit{do_min}(X)$ and $\mathit{do_max}(X)$, respectively.
Finally, \sys
appends two axioms and adds a single choice rule that allows the solver to select which properties to extract from the program in order to satisfy the query.
This consists as the core of the \sys logic program, and the reason the compilation of rules to $\mathsf{can}/\mathsf{do}$ predicates is required.
The next section describes the ASP programming paradigm.

@@

\heading{Logic program resolution} 
The \sys prototype compiles each input query to an answer-set programming (ASP) program~\cite{Eiter_2009}.
ASP \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo}~\cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

Clingo's syntax is similar to \textit{Prolog} and \textit{Datalog}, with code and data being represented
through logical terms.
Collections of logical terms, as seen in
\cref{lst:terms}, can represent the program and its properties.
Rules, expressed using
the \ttt{:-} operator, define relationships
between atoms.
Choice rules, exemplified in \cref{lst:choice}, allow the ASP
solver to make selections among atoms.
Integrity constraints, as shown in
\cref{lst:constraint}, restrict invalid answer sets.
Optimization
directives, such as \ttt{\#minimize} and \ttt{\#maximize} in
\cref{lst:optimization}, guide the solver to output optimal answer sets,
with respect to a given metric (\eg the number of properties to be extracted from the input program).

\heading{Logic space to concrete space}
Once the input query is compiled into an ASP program, \sys invokes the Clingo
ASP solver to compute a stable model that satisfies the query.
The model consists of a set of selected $\mathit{do}$ atoms representing the regeneration plan.
\sys then executes this plan~\cref{alg:resolution}.
First, \sys discards all plugin-activation atoms from the plan, 
as they are only relevant to the logic engine.
Then, \sys iterates through each
$\mathit{do}(f(a_1,\dots,a_n))$ atom.
If an argument value $a_i$ is missing, the plugin responsible for producing
it (found by parsing the variable's name) is executed and the produced program properties are stored in a value map.
\sys operates at the boundary between two representational domains: a symbolic
\emph{logic space}, where transformations over program properties are expressed, and
a \emph{concrete space}, where actual programs, inputs, outputs, and traces
reside.
As shown in \cref{fig:logic-to-concrete}, this process constitutes a
translation barrier: declarative goals over properties like $I$ (inputs)
$O$ (outputs), and $T$ (traces) are encoded in the query by the user and realized by \sys plugins through observation and instrumentation.
Once all symbolic properties have been concretized in dependency order, the value map containing them is passed to the regeneration phase.

@@

% \heading{Logic and concrete space}
% In the logic space, properties are encoded as predicates over abstract program
% identifiers.
% To bridge this abstraction gap, \sys given the set of true predicates
% from the logic space, extracts the corresponding concrete properties from the
% source program.
% For example, the predicate $\ttt{inputs}(p, I)$ in the logic space
% will be concertized to set of inputs $I$ by invoking the \ttt{input} plugin,
% which uses an LLM to generate a set of inputs for the program $p$.
% Similarly, the predicate $\ttt{graph}(p, I, O)$ will be concertized to a set of
% input-output pairs $(I, O)$ by executing the program $p$ on the inputs $I$ and
% recording the outputs $O$.
% After regeneration, the predicates over the concrete regenerated program $p\_$
% will be checked regarding their relationship to the same properties over program $p$
% in the logic space.
% % An equality relationship is one where the predicate's grounded version
% % $X$ is the same between the two programs, modulo the program identifier~(\eg, $\ttt{graph}(p, i, o) \equiv \ttt{graph}(p\_, i, o)$).
% Equality is defined differently for each property, and is enforced by the corresponding plugin.
% Similarly, minimization and maximization relationships are defined as
% the concrete property value being less-than or greater-than the original program's value, respectively.
% Again, the less-than and greater-than operators are defined by the plugin that verifies the property.

@@

@@

@@

For instance, a plan for regenerating a specific function $f$ might include (here: $\mathsf{do}$ surrounding predicate omitted)
\ttt{language}(f, \ttt{haskell}) the function must be written in Haskell, 
\ttt{pure}(f)) the function must be side-effect-free, and 
\ttt{fsignt}(f, s)) the function must keep the original program's signature $s$.

@@

\section{Verification and Feedback}
\label{sec:verification}

@@

\begin{table}[htpb]
  \centering
  \caption{\textbf{Correctness result summary.} Number of tasks passing all tests (when present) and manual inspection.}
  \begin{tabular}{lcc}
    \toprule
    Benchmark & Correctness & Naive LLM \\
    \midrule
    Rosetta Code & 16/16 & 16/16  \\
    Short Utilities & 13/13 & 13/13 \\
    SSCA & 3/3 & 0/3 \\
    IOCCC & 1/2 & 0/2 \\
    Microbenchmarks & 2.5/3 & 1/3 \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\heading{Idea of program projection}
\sys can be viewed as a projection system, as shown in \cref{fig:projection}.
\sys projects a program into an abstract space of properties, 
then, it performs transformations within that space (the input query),
and finally, it regenerates this projection back into a concrete program.
The idea of casting programs into a high-dimensioanl space has been studied before~\cite{ben-nun2018neural,alon2019code2vec,ellis2020dreamcoder,nye2021platypus},
but \sys's perspective might offer interesting insights.
